\documentclass[acmsmall,10pt,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmYear{2022}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
% \startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

\input{preamble}

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
% \citestyle{acmauthoryear}

\begin{document}

\title{Dependent Session Types for Certified Concurrent Programming}

\author{Qiancheng Fu}
\affiliation{
  \institution{Boston University}
  \city{Boston}
  \state{MA}
  \country{USA}
}
\email{qcfu@bu.edu}

\author{Zachery Casey}
\affiliation{
  \institution{Boston University}
  \city{Boston}
  \state{MA}
  \country{USA}
}
\email{caseyz@bu.edu}

\author{Hongwei Xi}
\affiliation{
  \institution{Boston University}
  \city{Boston}
  \state{MA}
  \country{USA}
}
\email{hwxi@bu.edu}

\begin{abstract}
We present \TLLC{} which extends the Two-Level Linear dependent type theory
(TLL) with session type based concurrency. Equipped with Martin-L\"{o}f style
dependency, the session types of \TLLC{} allow protocols to specify the
properties of communicated messages. When used in conjunction with the dependent
type machinery already present in TLL, dependent session types facilitate the a
form of relational verification by relating concurrent programs with their
idealized sequential counterparts. Correctness properties proven for sequential
programs can now be easily lifted to their corresponding concurrent programs.
Session types now become a powerful tool for intrinsically verifying the
correctness of data structures such as queues and concurrent algorithms such as
map-reduce. To extend TLL with session types, we develop a novel formulation of
intuitionistic session type which we believe to be widely applicable for
integrating session types into other type systems beyond the context of \TLLC{}.
We study the meta-theory of our language, proving its soundness as both a term
calculus and a process calculus. All reported results are formalized in Coq.
A prototype compiler which compiles \TLLC{} programs into concurrent C code is
implemented and freely available.
\end{abstract}
\keywords{dependent types, linear types, session types, concurrency}

\maketitle

\section{Introduction}
\input{intro}

\section{Overview of Dependent Session Types}
\input{overview}

\section{Relational Verification via Dependent Session Types}
\input{relational}

\section{Formal Theory of Dependent Session Types}
\input{theory}

\bibliography{reference}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
